\vspace{-10mm}\bf \mytitle

### 2021-09-02 ms.Rnw
code to make unformatted accepted version

<<debug,echo=F>>=
run_level <- 1
#run_level <- 2
#run_level <- "jasa-sept-2020"
bm_run_level <- run_level
mscale_run_level <- run_level
slice_run_level <- run_level

@@

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

@@

\newcommand\rewrite[1]{{\textcolor{red}{#1}}}

\usepackage{fullpage}
\addtolength{\textheight}{-.3in}%

@@

\if1\blind
{
  \title{\vspace{-10mm}\bf \mytitle}
  \author{Edward L. Ionides\thanks{This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grants 1-U54-GM111274 and 1-U01-GM110712.}\hspace{.2cm}\\
    Department of Statistics, University of Michigan\\
    Kidus Asfaw\\
    Department of Statistics, University of Michigan\\
    %and \\
    Joonha Park\\
    Department of Mathematics \& Statistics, Boston University\\
    %and\\
    Aaron A. King\\
    Department of Ecology \& Evolutionary Biology, University of Michigan
}
  \maketitle
} \fi

@@

\begin{abstract}
Bagging (i.e., bootstrap aggregating) involves combining an ensemble of bootstrap estimators.
The particle filter, also known as sequential Monte Carlo or the bootstrap filter, is a widely used algorithm for estimating the latent states or likelihood of a partially observed Markov process.
Our bagged filter (BF) methodology combines an ensemble of particle filters, using spatiotemporally localized weights to select successful filters at each point of space and time.
In some situations, BF methodology can theoretically beat a curse of dimensionality affecting standard particle filters;
in others, BF fails in theory to beat the curse, but nevertheless evinces practical advantages.
Our focus is on evaluation of the likelihood function, though BF theory and methodology are also pertinent to latent state estimation.
Applications suited to BF methodology include spatiotemporal analysis of epidemiological and ecological systems.
We show that BF methodology can out-perform an ensemble Kalman filter on a coupled population dynamics model arising in the epidemiology of infectious disease.
We also find that a block particle filter performs well on this task, though the bagged filter respects smoothness and conservation laws that a block particle filter can violate.

@@

@@

Particle filter (PF) algorithms are a widespread tool for inference on nonlinear partially observed dynamic systems \citep{doucet01,doucet11,breto09,ionides15}.
Modern particle filtering traces back to \citet{gordon93} where the algorithm was called the {\it bootstrap filter}. 
Standard PF methods suffer from a curse of dimensionality (COD), defined as an exponential increase in computational demands as the problem size grows, which limits its applicability to spatiotemporal systems \citep{bengtsson08,snyder15,rebeschini15}.
The COD presents empirically as numerical instability of the Monte Carlo algorithm for attainable numbers of particles.
Bootstrap aggregating (i.e., bagging) is a methodology to improve numerically unstable estimators by combining an ensemble of replicated bootstrap calculations \citep{breiman96}.
We consider algorithms combining an ensemble of replicated particle filters, and we use the name {\it bagged filter}.
The task of each bootstrap replicate is to provide some predictive skill at some points in space and time.
The bagged filter then proceeds to combine these replicates, assessing which replicates are reliable predictors at a given spatiotemporal location.

Much previous research has investigated scalable approaches to filtering and inference with applications to spatiotemporal systems.
Our bagged filters are in the class of {\it plug-and-play} algorithms, meaning that they require as input a simulator for the latent dynamic process but not an evaluator of transition probabilities \citep{breto09,he10}.
This simulation-based approach, also known as {\it likelihood-free} \citep{brehmer20} or {\it equation-free} \citep{kevrekidis09}, facilitates application to a wide class of models.
The ensemble Kalman filter \citep{evensen96,katzfuss19} is a widely used plug-and-play method which uses simulations to parameterize a Gaussian-inspired filtering rule.
Another plug-and-play approach to combat the COD is the block particle filter \citep{rebeschini15,ng02}.
Both ensemble Kalman filter and block particle filter methods construct trajectories that can violate smoothness and conservation properties of the dynamic model.
By contrast, our bagged filters are built using valid trajectories of the dynamic model, but make localization approximations when comparing these trajectories to data.
It is an empirical question how these strengths and weaknesses play out for a particular data analysis.
The plug-and-play property of an algorithm makes a general numerical implementation possible: we provide a software environment within which alternative methods can be applied.

The replicated stochastic trajectories in a bagged filter form an ensemble of representations of the dynamic system.
Unlike ensemble Kalman filters \citep{lei10} the bagged replicates are independent in a Monte Carlo sense.
Bagged filters therefore bear some resemblance to \textit{poor man's ensemble} forecasting methodology in which a collection of independently constructed forecasts is generated using different models and methods \citep{ebert01}. 
Poor man's ensembles have sometimes been found to have greater forecasting skill than any one forecast \citep{leutbecher08,palmer02,chandler13}.
One explanation for this phenomenon is that even a hypothetically perfect model cannot filter and forecast well using methodology afflicted by the COD.  
We show that bagged filter methodology can relieve this limitation.
From this perspective, the independence of the forecasts in the poor man's ensemble, rather than the diversity of model structures, may be the key to its success.

In a basic bagged filter, each replicate simply simulates a realization of the latent process model, corresponding to a degenerate instance of PF with only a single particle.
We call this the unadapted bagged filter ({\UBF}) since each replicate in the ensemble does not track the data.
While formally beating the COD under a weak mixing assumption, {\UBF} can have poor numerical behavior if a very large number of replicates are needed to reach this happy asymptotic limit.
Our empirical results show that {\UBF} may indeed be a useful algorithm in some situations.
More importantly, however, it provides a route to two other algorithms, which we call the adapted bagged filter (\ABF) and adapted bagged filter with intermediate resampling (\ABFIR).
{\ABF} and {\ABFIR} carry out some resampling on each replicate and therefore enable each replicate to track the data, in a weak sense, while resisting the COD.
This comes at a price: {\ABF} no longer asymptotically beats COD, and {\ABFIR} does so only under restrictive conditions.
However, we demonstrate useful behavior for these algorithms on finite samples.

To set up notation, we suppose the latent process is comprised of a collection of units, labeled $\{1,2,\dots,\Unit\}$ which we write as $1\mycolon\Unit$. 
An unobserved, latent Markov process at a discrete sequence of observation times is written as $\{\myvec{X}_{\time},\time\in 0\mycolon\Time\}$, with $\myvec{X}_{\time}=X_{1:\Unit,\time}$ taking values in $\Xspace^\Unit$.
The initial value $\myvec{X}_{0}$ may be stochastic or deterministic.
Observations are made on each unit, modeled by an observable process $\{\myvec{Y}_{\time}=Y_{1:\Unit,\time},\time\in 1\mycolon\Time\}$ which takes values in $\Yspace^\Unit$.
Observations are modeled as being conditionally independent given the latent process.
The conditional independence of measurements applies over both space and time, so $Y_{\unit\comma\time}$ is conditionally independent of
$\big\{
  X_{\altUnit\comma\altTime},
  Y_{\altUnit\comma\altTime},
  (\altUnit\comma\altTime)\neq(\unit\comma\time)
\big\}$
given $X_{\unit\comma\time}$.
This unit structure for the observation process is not necessary for all that follows (see Sec.~\SuppSecGeneralization).
We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  of $\{X_{1:\Unit,\time}\}$ and $\{Y_{1:\Unit,\time}\}$ with respect to some appropriate measure.
We use the same subscript notation to denote other marginal and conditional densities.
The data are ${y}_{\unit\comma\time}$ for spatial unit $\unit$ at time $\time$. 
For a fixed value of $\Unit$, this model fits into the partially observed Markov process (POMP) framework \citep{breto09}, also known as a state space model or hidden Markov model.
Spatiotemporal POMP models are characterized by a latent Markov process with dimension proportional to $\Unit$.
We use the term {SpatPOMP} to describe POMP models indexed by $\unit$ as well as $\time$.
In accordance with the spatial interpretation of the units $\unit\in 1\mycolon\Unit$, we may suppose that units sufficiently distant in some sense approach independence, a phenomenon called {\it weak coupling}. 
It is worth pointing out that inference techniques designed for SpatPOMPs are also relevant in high-dimensional stochastic dynamic systems where $\unit$ does not index spatial locations.
In the following, we use the ordering on the set of observations corresponding to their labels $1\mycolon\Unit$ to define the set of observations preceding unit $\unit$ at time $\time$ as
\begin{equation}\label{eq:setA}
A_{\unit\comma\time}=\big\{(\tilde\unit,\tilde\time): 1 \leq \tilde\time<\time \mbox{ or } (\tilde\time=\time \mbox{ and } \tilde\unit <\unit)\big\}.
\end{equation}
The ordering of the spatial locations in \myeqref{eq:setA} might seem artificial, and indeed densities such as
$f_{X_{\unit\comma\time}|X_{A_{\unit\comma\time}}}$ will frequently be hard to compute or simulate from.
However, the algorithms we study do not carry out such computations, but only evaluate the measurement model on these neighborhoods.
To represent weak coupling, we suppose there is a neighborhood $B_{\unit\comma\time}\subset A_{\unit\comma\time}$ such that the latent process on 
$A_{\unit\comma\time} \setminus B_{\unit\comma\time}$ 
is approximately conditionally independent of $X_{\unit\comma\time}$ given data on $B_{\unit\comma\time}$.

The inputs for the {\UBF}, {\ABF} and {\ABFIR} are listed in Table~1.
The plug-and-play property is evident since all three algorithms require as input a simulator for the latent dynamic process but not an evaluator of transition probabilities.
Our primary interest is in the estimation of the log likelihood for the data given the model, $\loglik=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time})$, which is of fundamental importance in both Bayesian and non-Bayesian statistical inference.
Monte Carlo estimates of the log likelihood underpin modern methods for full-information statistical inference on partially observed dynamic systems \citep{andrieu10,ionides15,ionides17}.
Filtering theory ties together the three tasks of likelihood evaluation, forecasting, and inferring latent dynamic variables \citep{crisan11}.
Successful likelihood evaluation is therefore closely related to success at forecasting and latent process reconstruction.

The pseudocodes for the algorithms that follow will adopt a convention that implicit loops are carried out over all free indices.
For example, the construction of $w^P_{\unit,\time,\rep}$ in {\UBF} has an implicit loop over $\unit$, $\time$ and $\rep$. 
The construction of $w^P_{\unit,\time,\rep,\np}$ in {\ABF} has an implicit loop over $\unit$, $\rep$ and $\np$ but not over $\time$ since $\time$ is already defined in an explicit loop and is therefore not a free index.
Table~1 lists the ranges of the implicit loops for all following algorithms.

\begin{center}
\noindent\begin{tabular}{l}
\hline
\inputSpace {\bf Table 1. Bagged filter inputs, outputs and implicit loops.}\\
\hline
\inputSpace {\bf input:}
\\
simulator for $f_{\myvec{X}_0}(\myvec{x}_0)$ and $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$\\
    evaluator for $f_{{Y}_{\unit\comma\time}|{X}_{\unit\comma\time}}({y}_{\unit\comma\time}\given {x}_{\unit\comma\time})$\\
    number of replicates, ${\Rep}$\\
    neighborhood structure, $B_{\unit\comma\time}$\\
    data, $\data{\myvec{y}}_{1:\Time}$\\
{\ABF} and {\ABFIR}:    particles per replicate,  $\Np$\\
{\ABFIR}: number of intermediate timesteps, $\Ninter$ \\
{\ABFIR}: measurement variance parameterizations, ${\VtoTheta}_{\unit\comma\time}$ and ${\thetaToV}_{\unit\comma\time}$\\
{\ABFIR}: approximate process and observation mean functions, $\myvec{\mu}$ and $h_{\unit\comma\time}$\\
\inputSpace {\bf output:}\\
  Log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
\inputSpace {\bf implicit loops:}\\
$\; \unit \mbox{ in } \seq{1}{\Unit}$, 
$\; \time \mbox{ in } \seq{1}{\Time}$, 
$\; \rep \mbox{ in } \seq{1}{\Rep}$, 
$\; \np \mbox{ in } \seq{1}{\Np}$
%$\; \npgir \mbox{ in } \seq{1}{\Npgir}$
\lastLineSpace \\
\hline
\end{tabular}
\end{center}

The~{\UBF} algorithm that follows amounts to a bagged filter with a single particle per replicate, with no weighting or resampling within each replicate.

%%%%%%%%%%%%% ttttttttttttttttttttttttt

\begin{center}
\noindent\begin{tabular}{l}
\hline
\inputSpace {\bf {\UBF}. Unadapted bagged filter.}\\
\hline
%\inputSpace {\bf input:} From Table~1, with $\Np=1$ \\
%\hline
\firstLineSpace 
Simulate $\myvec{X}^{\tif}_{0:\Time,\rep}\sim f_{\myvec{X}_{0:\Time}}(\myvec{x}_{0:\Time})$\\ 
Measurement weights,
$w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}\given X^{\tif}_{\unit,\time,\rep})$
\\
Prediction weights, 
$w^P_{\unit,\time,\rep}=\prod_{(\tilde \unit,\tilde n)\in B_{\unit,\time}}
w^M_{\tilde\unit,\tilde n,\rep}$\\
$\MC{\loglik}_{\unit,\time}= 
\log\left(
  \sum_{\rep=1}^\Rep w^M_{\unit,\time,\rep}w^P_{\unit,\time,\rep}
\right)
-\log\left(
  \sum_{\rep=1}^\Rep w^P_{\unit,\time,\rep}
\right)
$
\lastLineSpace
\\
\hline
\end{tabular}
\end{center}

Theorem~\ref{thm:tif} shows that {\UBF} can beat COD. 
However, an algorithm such as {\UBF}, which includes no resampling, will fail on long time series absent rapid spatiotemporal mixing, which allows simulated particles to remain relevant over the course of a long time series.
It may in practice be necessary to select simulations consistent with the data, much as standard PF algorithms do.
We look for approaches that build on the basic insight of {\UBF} while having superior practical performance.

Whereas the full global filtering problem may be intractable via importance sampling methods, a version of this problem local in space and time may nevertheless be feasible.
Drawing a sample from the conditional density, $f_{\myvec{X}_{\time}|\myvec{Y}_{\time},\myvec{X}_{\time-1}}$, is distinct from the filtering problem: we call this operation {\it adapted simulation}.
For models where $\myvec{X}_{\time}$ is close to $\myvec{X}_{\time-1}$, adapted simulation requires a local calculation that may be much easier than the full filtering calculation.
The following adapted bagged filter ({\ABF}) is constructed under a hypothesis that the adapted simulation problem is tractable.
The adapted simulations are then reweighted in a neighborhood of each point in space and time to construct a local approximation to the filtering problem which leads to an estimate of the likelihood.

@@

\section{Likelihood factorizations and their approximations}

>>>>

<<debug,echo=F>>=
# run_level <- "jasa-sept-2020"
 run_level <- "manual"
# run_level <- 2

if(is.numeric(run_level)){
  bm_run_level <- run_level
  mscale_run_level <- run_level
  slice_run_level <- run_level
}

if(run_level=="manual"){
  bm_run_level <- 4
  mscale_run_level <- 5
  slice_run_level <- 3
}

@@

\usepackage{amsmath,amssymb}
%\usepackage{graphicx,psfrag,epsf}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

@@

\newcommand\rewrite[1]{{\textcolor{red}{#1}}}
%\newcommand\jpc[1]{{\textcolor{orange}{#1}}}
\newcommand\aak[1]{{\textcolor{blue}{#1}}}

\usepackage{fullpage}
% DON'T change margins - should be 1 inch all around.
%\addtolength{\oddsidemargin}{-.5in}%
%\addtolength{\evensidemargin}{-.5in}%
%\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
%%\addtolength{\textheight}{.3in}%
%\addtolength{\topmargin}{-.8in}%

@@

\if1\blind
{
  \title{\vspace{-10mm}\bf \mytitle}
  \author{Edward L. Ionides\thanks{This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grants 1-U54-GM111274, 1-U01-GM110712 and 1-R01-AI143852. We acknowledge constructive feedback from two anonymous referees and the associate editor.}\hspace{.2cm}\\
    Department of Statistics, University of Michigan\\
    Kidus Asfaw\\
    Department of Statistics, University of Michigan\\
    %and \\
    Joonha Park\\
    Department of Mathematics, University of Kansas\\
    %and\\
    Aaron A. King\\
    Department of Ecology and Evolutionary Biology \& \\ Center for the Study of Complex Systems, University of Michigan
}
  \maketitle
} \fi

@@

\begin{abstract}
Bagging (i.e., bootstrap aggregating) involves combining an ensemble of bootstrap estimators.
We consider bagging for inference from noisy or incomplete measurements on a collection of interacting stochastic dynamic systems.
Each system is called a unit, and each unit is associated with a spatial location.
A motivating example arises in epidemiology, where each unit is a city: the majority of transmission occurs within a city, with smaller yet epidemiologically important interactions arising from disease transmission between cities.
Monte~Carlo filtering methods used for inference on nonlinear non-Gaussian systems can suffer from a curse of dimensionality as the number of units increases.
We introduce bagged filter (BF) methodology which combines an ensemble of Monte Carlo filters, using spatiotemporally localized weights to select successful filters at each unit and time.
We obtain conditions under which likelihood evaluation using a BF algorithm can beat a curse of dimensionality, and we demonstrate applicability even when these conditions do not hold.
BF can out-perform an ensemble Kalman filter on a coupled population dynamics model describing infectious disease transmission.
A block particle filter also performs well on this task, though the bagged filter respects smoothness and conservation laws that a block particle filter can violate.

@@

%%% JASA instructions
%% Should be written with the following elements in the following order: title page; author footnote; abstract; keywords; article text (table(s); figures); acknowledgments; appendices; references
%% Should be no more than 35 pages, inclusive of the abstract, tables, references, figure captions, footnotes, endnotes.
%% Should contain an unstructured abstract of 200 words.
%% Should contain between 3 and 5 keywords. Read making your article more discoverable, including information on choosing a title and search engine optimization.

@@

Bagging is a technique to improve numerically unstable estimators by combining an ensemble of replicated bootstrap calculations \citep{breiman96}.
In the context of nonlinear partially observed dynamic systems, the bootstrap filter of \citet{gordon93} has led to a variety of particle filter (PF) methodologies \citep{doucet01,doucet11};
Here, we consider algorithms combining an ensemble of replicated particle filters, which we term {\it bagged filter} algorithms.
Standard PF methods suffer from a curse of dimensionality (COD), defined as an exponential increase in computational requirement as the problem size grows, limiting its applicability to large systems \citep{bengtsson08,snyder15,rebeschini15}.
The COD presents empirically as numerical instability of the Monte Carlo algorithm for affordable numbers of particles.
Much previous research has investigated scalable approaches to filtering and inference with applications to spatiotemporal systems.
Our bagged filters are in the class of {\it plug-and-play} algorithms, meaning that they require as input a simulator for the latent dynamic process but not an evaluator of transition probabilities \citep{breto09,he10}.
Similar properties to plug-and-play are {\it likelihood-free} \citep{brehmer20} and {\it equation-free} \citep{kevrekidis09}.
The ensemble Kalman filter \citep{evensen09book,lei10,katzfuss19} is a widely used plug-and-play method which uses simulations to construct a nonlinear filter that is exact for a linear Gaussian model.
Another plug-and-play approach to combat the COD is the block particle filter \citep{rebeschini15,ng02}.
Both ensemble Kalman filter and block particle filter methods construct trajectories that can violate smoothness and conservation properties of the dynamic model.
By contrast, our bagged filters are built using valid trajectories of the dynamic model, making localization approximations only when comparing these trajectories to data.

The replicated stochastic trajectories in a bagged filter form an ensemble of representations of the dynamic system.
Unlike the particles in a particle filter or ensemble Kalman filter, the bagged replicates are independent in a Monte Carlo sense.
Bagged filters therefore bear some resemblance to \textit{poor man's ensemble} forecasting methodology in which a collection of independently constructed forecasts is generated using different models and methods \citep{ebert01}. 
Poor man's ensembles have sometimes been found to have greater forecasting skill than any one forecast \citep{leutbecher08,palmer02,chandler13}.
One explanation for this phenomenon is that even a hypothetically perfect model cannot provide effective filtering using methodology afflicted by the COD.  
We show that bagged filter methodology can relieve this limitation.
From this perspective, the independence of the forecasts in the poor man's ensemble, rather than the diversity of model structures, may be the key to its success.

We first consider a simple bagged filter where each replicate is an independent simulation of the latent process model.
We call this the unadapted bagged filter ({\UBF}) since the replicates in the ensemble depend on the model but not on the data.
{\UBF} is described in Sec.~\ref{sec:ubf}, with a theoretical analysis presented in Sec.~\ref{sec:ubf:theory}.
Each {\UBF} replicate corresponds to a basic PF algorithm with a single particle.
We show that {\UBF} formally beats the COD under a weak mixing assumption, though {\UBF} can have poor numerical behavior if a very large number of replicates are needed to reach this happy asymptotic limit.
Subsequent empirical results show that {\UBF} may nevertheless be a useful algorithm in some situations.
In Sec.~\ref{sec:abf}, we generalize {\UBF} to construct an adapted bagged filter (\ABF) where each replicate tracks the data.
The price of adaptation is that {\ABF} no longer avoids the COD, a limitation that can be controlled in certain situations by supplementing {\ABF} with a technique called intermediate resampling, to obtain the {\ABFIR} algorithm.
Theoretical results for {\ABF} and {\ABFIR} algorithms are developed in Sec.~\ref{sec:abf:theory}.
The algorithms are demonstrated in action and compared with alternative approaches in Sec.~\ref{sec:examples}.

\section{The unadapted bagged filter (UBF)}
\label{sec:ubf}

Suppose the collection of units is indexed by the set $\{1,2,\dots,\Unit\}$, which is written as $1\mycolon\Unit$.
The latent Markov process is denoted by $\{\myvec{X}_{\time},\time\in 0\mycolon\Time\}$, with $\myvec{X}_{\time}=X_{1:\Unit,\time}$ taking values in a product space $\Xspace^\Unit$.
This discrete time process may arise from a continuous time Markov process $\{\myvec{X}(t), t_0\le t\le t_{\Time} \}$ observed at times $t_{1:\Time}$, and in this case we set $\myvec{X}_\time=\myvec{X}(t_\time)$.
The initial value $\myvec{X}_{0}$ may be stochastic or deterministic.
Observations are made on each unit, modeled by an observable process $\{\myvec{Y}_{\time}=Y_{1:\Unit,\time},\time\in 1\mycolon\Time\}$ which takes values in a product space $\Yspace^\Unit$.
Observations are modeled as being conditionally independent given the latent process.
The conditional independence of measurements applies over both time and the unit structure, so the collection $\big\{Y_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$ is conditionally independent given
$\big\{X_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$.
The unit structure for the observation process is not necessary for all that follows (see Sec.~\SuppSecGeneralization).
We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  of $X_{1:\Unit,0:\Time}$ and $Y_{1:\Unit,1:\Time}$ with respect to some appropriate measure, following a notational convention that the subscripts of $f$ denote the joint or conditional density under consideration.
The data are $\data{y}_{\unit\comma\time}$ for unit $\unit$ at time $\time$.
This model is a special case of a  partially observed Markov process  \citep[POMP,][]{breto09}, also known as a state space model or hidden Markov model.
The additional unit structure, not generally required for a POMP, is appropriate for modeling interactions between units characterized by a spatial location, and so we call the model a SpatPOMP.
In the following, we use a lexicographical ordering on the set of observations;
Specifically, we define the set of observations preceding unit $\unit$ at time $\time$ as
\begin{equation}\label{eq:setA}
A_{\unit\comma\time}=\big\{(\tilde\unit,\tilde\time): 1 \leq \tilde\time<\time \mbox{ or } (\tilde\time=\time \mbox{ and } \tilde\unit <\unit)\big\}.
\end{equation}
The ordering of the spatial locations in \myeqref{eq:setA} might seem artificial, and indeed densities such as
$f_{X_{\unit\comma\time}|X_{A_{\unit\comma\time}}}$ will frequently be hard to compute or simulate from.
The bagged filter algorithms we study do not evaluate or simulate such transition densities but only compute the measurement model on neighborhoods, unlike the filter of \citet{beskos17} built on a similar factorization. 
If sufficiently distant units are approximately independent, we say the system is {\it weakly coupled}.
In this case, we suppose there is a neighborhood $B_{\unit\comma\time}\subset A_{\unit\comma\time}$ such that the latent process on 
$A_{\unit\comma\time} \setminus B_{\unit\comma\time}$ 
is approximately conditionally independent of $X_{\unit\comma\time}$ given data on $B_{\unit\comma\time}$.

Our primary interest is estimation of the log likelihood for the data given the model, $\loglik=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time})$, which is of fundamental importance in both Bayesian and non-Bayesian statistical inference.
A general filtering problem is to evaluate $\E\big[h(X_{\unit,\time}) \given Y_{A_{\unit,\time}}\equals \data{y}_{A_{\unit,\time}}\big]$ for some function $h:\Xspace\to\R$.
Taking $h(x)= f_{Y_{\unit,\time}| X_{\unit,\time}}\big(\data{y}_{\unit,\time}\given x\big)$ gives a filtering representation of the likelihood evaluation problem.
Further discussion on bagged filtering for other filtering problems is given in Sec.~{\SuppSecLatentStates}.
For likelihood-based inference, maximization plays an important role in point estimation, confidence interval construction, hypothesis testing and model selection. 
An extension of bagged filtering to likelihood maximization is demonstrated in Sec.~\ref{sec:profile} following the approach described in Sec.~{\SuppSecParameterEst}.

Pseudocode for a {\UBF} algorithm for likelihood evaluation is given below.
The prediction weight  $w^P_{u,n,i}$ gives an appropriate weighting for replicate $i$ for predicting $\data{y}_{u,n}$ based on the most relevant data, $\data{y}_{B_{u,n}}$.
Conditional log likelihoods are estimated using an approximation
\begin{eqnarray*}
    \ell_{u,n} &=& \log f_{Y_{u,n}|Y_{A_{u,n}}}\big(\data{y}_{u,n}\given \data{y}_{A_{u,n}}\big)
    = \log \left( \int f_{Y_{u,n}|X_{u,n}}(\data{y}_{u,n}\given x) \, f_{X_{u,n}|Y_{A_{u,n}}}(x \given \data{y}_{A_{u,n}}) \, dx \right)
\\
  &\approx& \log \left( \int f_{Y_{u,n}|X_{u,n}}(\data{y}_{u,n}\given x) \, f_{X_{u,n}|Y_{B_{u,n}}}(x\given \data{y}_{B_{u,n}}) \, dx \right).
\end{eqnarray*}
The choice of $B_{u,n}$ is determined empirically, with a bias-variance trade-off used to compare small neighborhoods such as $B_{u,n} = \{ (u,n-1),(u-1,n)\}$ or $B_{u,n} = \{ (u,n-1),(u,n-2)\}$ against larger neighborhoods.
The plug-and-play property is evident because {\UBF} requires as input a simulator for the latent coupled dynamic process but not an evaluator of transition probabilities.

%%%%%%%%%%%%% uuuuuuuuu

% UBF PSEUDOCODE uuuuuuuuuu
\begin{algorithm}[H]
  \caption{\bf Unadapted bagged filter (UBF).
   }\label{alg:ubf}
  \KwIn{
    simulator for $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$ and $f_{\myvec{X}_0}(\myvec{x}_0)$;
    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time})$;
    data, $\data{\myvec{y}}_{1:\Time}$;
    number of replicates, $\Rep$;
    neighborhood structure, $B_{\unit,\time}$
  }
\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
initialize simulation, $\myvec{X}_{0,\rep} \sim f_{\myvec{X}_0}(\cdot)$
  \;
  \nllabel{alg:ubf:for:n}
\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
simulate,
    $\myvec{X}_{\time,\rep} \sim
      f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}
      \big( \mydot \given \myvec{X}_{\time-1,\rep}
    \big)$
  \nllabel{alg:ubf:adapted:proposals}
  \;
measurement weights,
  $w^M_{\unit,\tilde n,\rep}=
    f_{Y_{\unit,\time}|X_{\unit,\time}}
    \big (\data{y}_{\unit,\time}\given {X}_{\unit,\time,\rep}
  \big)$
    for $\unit$ in $\seq{1}{\Unit}$
  \;
prediction weights,
  $w^P_{\unit,\time,\rep}=\prod_{(\tilde \unit,\tilde n)\in B_{\unit,\time}} w^M_{\tilde\unit,\tilde n,\rep}$
    for $\unit$ in $\seq{1}{\Unit}$
      \nllabel{alg:ubf:adapted:weights}
  \;
  }
}
%conditional log likelihood,
  $\MC{\loglik}_{\unit,\time}= 
    \log\left(
      \sum_{\rep=1}^\Rep w^M_{\unit,\time,\rep}w^P_{\unit,\time,\rep}
    \right)
    -\log\left(
      \sum_{\rep=1}^\Rep w^P_{\unit,\time,\rep}
    \right)$
  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$
  \;
\KwOut{
log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
}
\end{algorithm}

\subsection{{\UBF} theory}
\label{sec:ubf:theory}

A dataset $\data{\myvec{y}}_{1:\Time}$ with $\Unit$ units is modeled via a joint density  $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$.
We consider non-asymptotic bounds that apply for all values of $\Unit$ and $\Time$. 
To impose a requirement that distant regions of space-time behave similarly and have only weak dependence, we assert the following conditions which define constants $\eone$, $\etwo$ and $Q$ used to bound the bias and variance in Theorem~\ref{thm:tif}.
Stronger bounds are obtained when the conditions hold for small $\eone$, $\etwo$ and $Q$.

The two mixing conditions in Assumptions~\ref{A1} and~\ref{A:unconditional:mix} are subtly different. 
Assumption~\ref{A1} describes a conditional mixing property dependent on the data, whereas~\ref{A:unconditional:mix} asserts a form of unconditional mixing.
Although both capture a similar concept of weak coupling, conditional and unconditional mixing properties do not readily imply one another.
Assumption~\ref{A2} is a compactness condition of a type that has proved useful in the theory of particle filters despite the rarity of its holding exactly.
Theorem~\ref{thm:tif} shows that these conditions let {\UBF} compute the likelihood with a Monte Carlo variance of order $\Unit\Time\Rep^{-1}$ with a bias of order $\Unit\Time\epsilon$.

Theorem~\ref{thm:tif} does not guarantee uniformity over $\Unit$ and $\Time$ of the rate of convergence as $\Rep\to\infty$.
However, it does guarantee that the polynomial bounds in (\ref{th1:lik:bound}) and (\ref{th1:lik:bound2}) hold for sufficiently large $\Rep$.
The COD is characterized by exponential bounds, and so Theorem~\ref{thm:tif} shows a specific sense in which {\UBF} can avoid COD.
%% This formalizes the heuristic that {\UBF} uses prediction weights $w^P_{u,n,i}$ which scale exponentially with the size of local neighborhoods $B_{u,n}$, avoiding the use of weights that scale exponentially with increasing $U$.
Uniformity of the central limit convergence in Theorem~\ref{thm:tif} may be expected to hold via a Berry-Esseen theorem, but extension of existing Berry-Esseen results for dependent processes \citep{bentkus97,jirak16} is beyond the scope of this article.

The approximation error for {\UBF} can be divided into two sources: a localization bias due to conditioning on a finite neighborhood, and Monte Carlo error. 
The localization bias does not disappear in the limit as Monte Carlo effort increases.
It does become small as the conditioning neighborhood increases, but the Monte Carlo effort grows exponentially in the size of this neighborhood.
Although the filtering inference is carried out using localization, the simulation of the process is carried out globally which avoids the introduction of additional boundary effects and ensures that the simulations comply with any constraints satisfied by the model for the latent process.

\section{Adaptation and intermediate resampling}
\label{sec:abf}

Theorem~\ref{thm:tif} shows that {\UBF} can beat COD.
However, {\UBF} can perform poorly on long time series unless weak temporal dependence allows simulated sample paths to remain relevant over the course of a long time series. 
For example, we will find that {\UBF} performs well on an epidemiological model (Sec.~\ref{sec:discussion}) but less well on a geophysical model (Sec.~{\SuppSecLorenz}).
It is sometimes necessary to select simulations consistent with the data, much as standard PF algorithms do.
We look for approaches that build on the basic insight of {\UBF} while having superior practical performance.

Whereas the full global filtering problem of drawing from $f_{\myvec{X}_{\time}|\myvec{Y}_{1:\time}}$ may be intractable via importance sampling methods, a version of this problem localized in space and time may nevertheless be feasible.
The conditional density, $f_{\myvec{X}_{\time}|\myvec{Y}_{\time},\myvec{X}_{\time-1}}$, is called the {\it adapted density}, and simulating from this density is called {\it adapted simulation}.
For models where $\myvec{X}_{\time-1}$ is highly informative about $\myvec{X}_{\time}$, importance sampling for adapted simulation may be much easier than the full filter calculation.
The following adapted bagged filter ({\ABF}) is constructed under a hypothesis that the adapted simulation problem is tractable, and it is applicable when the number of units is prohibitive for Monte Carlo sampling from the full filter distribution but not for sampling from the adapted distribution.
In {\ABF}, the adapted simulations are reweighted in a neighborhood of each unit and time point to construct a local approximation to the filtering problem which leads to an estimate of the likelihood.
The pseudocode for {\ABF}, below, reduces to {\UBF} when using a single particle per repicate, $\Np=1$.

@@

%\section{Likelihood factorizations and their approximations}

\subsection{{\ABFIR} theory}
\label{sec:abf:theory}

We  start by considering a deterministic limit for infinite Monte Carlo effort and explaining why the {\ABF} and {\ABFIR} algorithms approximately target the likelihood function, subject to suitable mixing behavior.
